Temperature and wave exposure as drivers of morphology in two kelp species

### 2019-08-01 morphology_paper.Rmd
fixed issue when creating abiotic dataset. More writing added in results section. RDA test output html

---
title: Temperature and wave exposure as drivers of morphology in two kelp species
  along the coast of South Africa
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  html_notebook:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
  pdf_document:
    fig_caption: yes
    fig_crop: no
    fig_height: 7
    fig_width: 7
  html_document:
    df_print: paged
  word_document:
    reference_docx: morphology_paper_files/frontiers_template.docx
papersize: A4
tables: yes
fontsize: 10pt
---

@@

# Ecklonia
eck_data <- read.csv("data/morph_data/eck_morph_data.csv", sep = ";") %>%
  filter(site!= "Rocky Bank") %>%   # Not using Rocky Bank in this study
  filter(site!= "De Hoop")

# Laminaria
lam_data <- read.csv("data/morph_data/lam_morph_data.csv", sep = ";") %>%
  mutate(site = replace(site, site == "Buffels Bay", "Buffelsbaai")) %>%
  # NB: Even though I am changing the name, there will still be no lon/lat values for this site
  # We need to decide how we want to handle this. Give it the same lon/lat as Buffelsbaai?
  mutate(site = replace(site, site == "Buffels Bay South", "Buffelsbaai South")) %>%
  # NB: This may also not be an appropriate change
  mutate(site = replace(site, site == "Bordjies North", "Bordjies Deep")) %>%
  # Not using Rocky Bank in this study
  filter(site != "Rocky Bank")

@@

```{r Temperature variable processing, message=FALSE, warning=FALSE, include=FALSE}
### Plots for temperature variables 
positions_temp <- c("Betty's Bay", "Roman Rock","Kalk Bay", "A-Frame", "Miller's Point", "Baboon Rock",  "Batsata Rock", "Buffelsbaai","Buffelsbaai South","Olifantsbos","Scarborough", "Soetwater", "Kommetjie","Hout Bay","Oudekraal","Sea Point","St Helena Bay")

@@

# The site list key
wave_sites <- read_csv2("metadata/site_meta_complete.csv") %>% 
  select(site, wave_7, wave_15)

# Annual clims
wave_ann <- wave_model %>% 
  mutate(date = lubridate::month(date, label = TRUE)) %>% 
  filter(site != "FB025") %>% # FB_25 has issues with 'hs' and 'tp'
  group_by(site, date) %>% 
  summarise_all(funs(mean = mean, sd = sd, min = min, max = max), na.rm = T) %>%
  ungroup() %>% 
  select(-date) %>% 
  group_by(site) %>% 
  summarise_all(funs(mean), na.rm = T) %>% 
  mutate(date = "Ann") %>% 
  select(site, date, everything()) %>% 
  group_by(site) %>% 
  left_join(wave_sites, by = c("site" = "wave_7")) %>% 
  ungroup() %>% 
  select(-site, wave_15) %>%
  rename(site = site.y) %>% 
  select(site, everything()) %>% 
  na.omit()

wave_ann <- wave_ann[-c(9,15,20,22), -c(2,23)]

#remove useless coloumns
wave_ann <- wave_ann %>% 
  select(site, hs_mean, hs_min, hs_max, hs_sd, tp_mean, tp_min, tp_max, tp_sd) 
```

@@

```{r wave power map, echo=FALSE, fig.cap='Schematic of *E. maxima* on the left and *L. pallida* on the right. (Dyer 2018)'}
library(png)
library(grid)

img_wavepower <- readPNG("figures/CapePen_Total.PNG")
grid.raster(img_wavepower)
```

@@

Significant differences were found between cool temperate and warm temperate regimes for all *L. pallida* morphological differences with lamina length, number of digits and thallus mass the exceptions. Certian morphological characteristics such as stipe mass, stipe length, total length and stipe diameter had significanlty higher means for the cool temperate regimes when compared to the warm temperate regime. Conversely, lamina weight and lamina thickness has significantly lower means for the cool temperate regime compared to the warm temperate regime. When *L. pallida* morphological characteristics were grouped to wave exposure categories kruskal-wallis tests revealed significant differences amoungst categories for all morphological characteristics with lamina length, number of digits and thallus mass the exceptions. The morphological characteristics that were significantly different amoungst sites exibited similar patterns, with increasing mean values from the full sheltered to extremely sheltered categories and then a decrease in mean values for the remainder of the categories. Two morphological characteristics, namely lamina thickness and stipe diameter, both exibit higher variations for the exposed categories.

@@

All deep *E. maxima* morphological characteristics showed differences between cool temperate and warm temperate regimes except for frond mass, frond length and stipe mass. Mean values for primary length, primary width and stipe circumference were all significantly higher for the cool temperate regime than the warm temperate regime. The remaining morphological characteristics which showed significant differences between regimes exhibited significanlty lower values for the cool temperate regime compared to the warm temperate regime. Significant differences amoungst wave exposure categories were found for all morphological characteristics of deep *E. maxima* populations. Certian morphological characteristics showed a gradual increase in mean value with inncreasing wave exposure. The remaining morphological characteristics exhibited decreased values for only the higher exposure categories and then a sharp increase in value for the highest wave exposure category. In all cases, mean values for the highest wave exposure level (very exposed) was greatest compared to other levels of exposure.

@@

@@

```{r shallow ecklonia temp plots, echo=FALSE, message=FALSE, warning=FALSE, fig.cap ='Boxplot of morphological characteristics for shallow population *E. maxima* grouped by temperature regime.'}
ggarrange(EST1,EST2,EST3,EST4,EST5,EST6,EST7,EST8,EST9,EST10, labels = "AUTO", common.legend = FALSE, legend = "bottom")
```

```{r echo=FALSE, warning=FALSE, warning=FALSE, fig.cap ='Boxplot of morphological characteristics for shallow population *E. maxima* grouped by wave exposure.'}
ggarrange(ESW1,ESW2,ESW3,ESW4,ESW5,ESW6,ESW7,ESW8, ESW9, labels = "AUTO", common.legend = FALSE, legend = "bottom")
```

```{r rda analysis lamanaria, eval=FALSE, include=FALSE}

# remove unnecessary columns
lam_df <- subset(lam_data, select = -c(1,2,6))

#calculate mean values for lamanaria morphology
lam_mean <- lam_df %>% 
  subset(select = -c(1,2)) %>% 
  group_by(site) %>% 
  summarise_all(funs(mean), na.rm = T) %>% 
  filter(site != "Roman Rock") %>% 
  filter(site != "Miller's Point") %>%  #the scipt keeps dropping it when combing the site list with the metadata
  ungroup()

@@

# merge temp and wave data to create environmental data-frame
abio_clim <- merge(temp_ann, wave_ann, by = "site")

@@

#remove date column and extra Betty's Bay
abio_lam <- abio_lam[-c(5) , -c(2)]

@@

abio_lam <- cbind(lam_sites, abio_lam)

abio_lam <- abio_lam %>%
  remove_rownames %>%
  column_to_rownames(var = "site")

#run RDA for abiotic variables
rda_all_lam <- rda(bio_lam ~ ., data = abio_lam)

@@

# Parsimonious RDA
rda_lam <- rda(bio_lam ~ tp_sd + max_temp + hs_mean, data = abio_lam)
summary(rda_lam)

@@

lam_meta_rda <- merge(lam_sites, kelp_sites, by = "site")
lam_meta_rda <- lam_meta_rda[-c(1,9,11,14), ]

# Plot RDA
lam_rda <- ggord(rda_lam, lam_meta_rda$region, ellipse = TRUE, hull = FALSE, grp_title = "Region", ptslab = TRUE, parse = TRUE, addsize = 3.5, ext = 0.9) 
```

@@

```{r rda plots, eval=FALSE, include=FALSE}
ggarrange(lam_rda, rda_eck_deep,rda_eck_shallow, common.legend = TRUE, labels = "AUTO", legend = "bottom")
```

>>>>

---
title: Temperature and wave exposure as drivers of morphology in two kelp species
  along the coast of South Africa
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
geometry: margin=1in
output:
  html_document:
    df_print: paged
  pdf_document:
    fig_caption: yes
    fig_crop: no
    fig_height: 7
    fig_width: 7
  word_document:
    reference_docx: morphology_paper_files/frontiers_template.docx
  html_notebook:
    fig_caption: yes
    fig_height: 7
    fig_width: 7
papersize: A4
tables: yes
fontsize: 10pt
---

@@

# Ecklonia
eck_data <- read.csv("data/morph_data/eck_morph_data.csv", sep = ";") %>%
  filter(site!= "Rocky Bank") %>%   # Not using Rocky Bank in this study
  filter(site!= "De Hoop")
eck_data$site <- as.character(eck_data$site)

# Laminaria
lam_data <- read.csv("data/morph_data/lam_morph_data.csv", sep = ";") %>%
  # NB: This may also not be an appropriate change
  # Not using Rocky Bank in this study
  filter(site != "Rocky Bank")
lam_data$site <- as.character(lam_data$site)

@@

```{r Temperature variable processing, message=FALSE, warning=FALSE, include=FALSE}
### Plots for temperature variables 
positions_temp <- c("Betty's Bay", "Roman Rock","Kalk Bay", "A-Frame", "Miller's Point", "Baboon Rock", "Batsata Rock","Bordjies Deep","Buffelsbaai","Buffelsbaai South","Olifantsbos","Scarborough", "Soetwater", "Kommetjie","Hout Bay","Oudekraal","Sea Point","St Helena Bay")

@@

# The site list key
wave_sites <- read_csv2("metadata/site_meta_complete.csv") %>% 
  select(site, wave_7, wave_15)
# Annual clims
wave_ann <- wave_model %>% 
  mutate(date = lubridate::month(date, label = TRUE)) %>% 
  filter(site != "FB025") %>% # FB_25 has issues with 'hs' and 'tp'
  group_by(site, date) %>% 
  summarise_all(funs(mean = mean, sd = sd, min = min, max = max), na.rm = T) %>%
  ungroup() %>% 
  select(-date) %>% 
  group_by(site) %>% 
  summarise_all(funs(mean), na.rm = T) %>% 
  mutate(date = "Ann") %>% 
  select(site, date, everything()) %>% 
  group_by(site) %>% 
  left_join(wave_sites, by = c("site" = "wave_7")) %>% 
  ungroup() %>% 
  select(-site, wave_15) %>%
  rename(site = site.y) %>% 
  select(site, everything()) %>% 
  na.omit()

#remove useless coloumns
wave_ann <- wave_ann[-c(9,15,20,22), -c(2,23)]

wave_ann <- wave_ann %>% 
  select(site, hs_mean, hs_min, hs_max, hs_sd, tp_mean, tp_min, tp_max, tp_sd)

# Subset directions out
dirs <- subset(wave_model, select = c(1, 2, 5, 6))

# Annual
dirs_ann <- dirs %>% 
  mutate(date = lubridate::month(date, label = TRUE)) %>% 
  group_by(site, date) %>% 
  summarise(dir_min = min(dir, na.rm = T),
            dir_max = max(dir, na.rm = T),
            dir_range = range(dir, na.rm = T)[2]-range(dir, na.rm = T)[1],
            dir_median = median(dir, na.rm = T), 
            dirw_min = min(dirw, na.rm = T),
            dirw_max = max(dirw, na.rm = T),
            dirw_range = range(dirw, na.rm = T)[2]-range(dirw, na.rm = T)[1],
            dirw_median = median(dirw, na.rm = T)) %>% 
  ungroup() %>% 
  select(-date) %>% 
  group_by(site) %>% 
  summarise_all(funs(mean(., na.rm = T))) %>% 
  mutate(date = "Ann") %>% 
  select(site, date, everything())

# combine all direction metrics
dirs_ann <- dirs_ann %>% 
  group_by(site) %>% 
  gather(key = statistic, value = value, 
         3:4) %>% 
  mutate(clim_stat = paste(date, statistic, sep = "_")) %>% 
  select(-date, -statistic) %>% 
  spread(key = clim_stat, value = value) %>% 
  left_join(wave_sites, by = c("site" = "wave_7")) %>% 
  ungroup() %>% 
  select(-site, wave_15) %>%
  rename(site = site.y) %>% 
  select(site, everything()) %>% 
  na.omit()

# Combine
#these will be used later for the redudancy analysis
clim <- merge(wave_ann, dirs_ann, by = "site")  
clim <- clim[ , -c(12,13,16,17,18)]
#remove duplicate sites
clim <- clim[-c(1,6,7,8,12,18,20), ]

# merge temp and wave data to create environmental data-frame
abio_clim <- merge(temp_ann, clim, by = "site")
```

@@

```{r wave power map, echo=FALSE, fig.cap='wave power maps'}
library(png)
library(grid)

img_wavepower <- readPNG("figures/CapePen_Total.PNG")
legend_wave_power <- readPNG("figures/CapePen_Legend.PNG")
grid.raster(img_wavepower,legend_wave_power)
```

@@

Significant differences were found between cool temperate and warm temperate regimes for all *L. pallida* morphological characteristics with lamina length, number of digits and thallus mass the exceptions. Certian morphological characteristics such as stipe mass, stipe length, total length and stipe diameter had significanlty higher means kelps in cool temperate regimes when compared to kelps for the warm temperate regime. Conversely, lamina weight and lamina thickness had significantly lower means for kelps in the cool temperate regime compared to kelps in the warm temperate regime. When *L. pallida* morphological characteristics were grouped to wave exposure categories kruskal-wallis tests revealed significant differences amoungst categories for all morphological characteristics with lamina length, number of digits and thallus mass the exceptions. The morphological characteristics that were significantly different amoungst sites exibited similar patterns, with increasing mean values from the full sheltered to extremely sheltered categories and a decrease in mean values for the remainder of the categories. Two morphological characteristics, namely lamina thickness and stipe diameter, both exibit higher variations for the exposed categories compared to other morphological characteristics.

@@

Six out of the ten deep *E. maxima* morphological characteristics showed differences between cool temperate and warm temperate regions except for frond mass, frond length, stipe mass and epiphyte length. Mean values for primary length, primary width and stipe circumference were significantly higher for kelps from cool temperate regions than for kelps from warm temperate regions. The remaining morphological characteristics, stipe length, number of tufts and total length had significanlty lower values for kelps from cool temperate compared to kelps from the warm temperate regime. Significant differences amoungst wave exposure categories were found for all morphological characteristics of deep *E. maxima* populations. Certian morphological characteristics showed a gradual increase in value with increasing degree of wave exposure. The remaining morphological characteristics exhibited decreased values for only the higher wave exposure categories and then a sharp increase in value for the highest wave exposure category.

@@

 ylim(0,30) +

@@

```{r shallow ecklonia temp plots, echo=FALSE, message=FALSE, warning=FALSE, fig.cap ='Boxplot of morphological characteristics for shallow population *E. maxima* grouped by temperature regime.',fig.height=8, fig.width=8, include=TRUE}
ggarrange(EST1,EST2,EST3,EST4,EST5,EST6,EST7,EST8,EST9,EST10, labels = "AUTO", common.legend = FALSE, legend = "bottom")
```

```{r echo=FALSE, warning=FALSE, warning=FALSE, fig.cap ='Boxplot of morphological characteristics for shallow population *E. maxima* grouped by wave exposure.',fig.height=8, fig.width=8, include=TRUE}
ggarrange(ESW1,ESW2,ESW3,ESW4,ESW5,ESW6,ESW7,ESW8, ESW9, labels = "AUTO", common.legend = TRUE, legend = "bottom")
```

```{r rda analysis lamanaria, eval=FALSE, include=TRUE}
#data processed in the temperature and wave sections will be used here eg. abio_clim
# remove unnecessary columns
lam_df <- subset(lam_data, select = -c(1,2,6))

#calculate mean values for lamanaria morphology
lam_mean <- lam_df %>% 
  subset(select = -c(1,2)) %>% 
  group_by(site) %>% 
  summarise_all(funs(mean), na.rm = T) %>% 
  filter(site != "Roman Rock") %>% 
  ungroup()

@@

@@

#remove date column 
abio_lam <- abio_lam[ , -c(2)]

@@

#have an issue with Miller's Point not merging so removing for now. creating temp site list for this
lam_sites_temp <- lam_sites[-c(9), ]
abio_lam <- cbind(lam_sites_temp, abio_lam)

abio_lam <- abio_lam %>%
  remove_rownames %>%
  column_to_rownames(var = "site")
#remove Millers from bio_lam to match up sites as I am still having issues
bio_lam <- bio_lam[-c(9), ]
#run RDA for abiotic variables
rda_all_lam <- rda(bio_lam ~ ., data = abio_lam)

@@

# Parsimonious RDA
rda_lam <- rda(bio_lam ~ tp_sd + hs_mean + mean_temp + dir_median, data = abio_lam)
summary(rda_lam)
anova.cca(rda_lam, step = 1000)
anova.cca(rda_lam, step = 1000, by = "axis")

@@

meta_all <- read.csv("metadata/site_meta_complete.csv", sep = ";")
meta_lam_rda <- merge(lam_sites, meta_all, by = "site")
meta_lam_rda <- lam_meta_rda[-c(9), ] #remove Miller's point as it is dropping during processing

# Plot RDA
lam_rda <- ggord(rda_lam, meta_lam_rda$region,ellipse = TRUE, hull = FALSE, grp_title = "Region", ptslab = TRUE, parse = TRUE, addsize = 3.5, ext = 0.9) 
```

@@

```{r rda plots, echo=FALSE, message=FALSE, warning=FALSE}
lam_rda
```
